Data Specialization for a Staged Lambda Calculus

### 2015-02-11 paper/example.tex
fixed figures
edited section 2.

@@

\crem{
Notes from 1/16:
\begin{itemize}
\item We are modeling computations where some inputs arrive at stage 1 and
others arrive at stage 2. As a result, some subcomputations can be performed at
stage 1, and others at stage 2.
\item \emph{Statics}: M12 is a language with three worlds: 1, 2, and M.
We model the computations described above as functions in world M.
\item \emph{Dynamics}: We describe how to perform all of the stage 1
subcomputations in an M12 program. Our semantics are mutually defined in terms
of how to reduce terms in each of the three worlds.
\end{itemize}
}

@@

Quickselect, which we define in \ref{fig:qs-unstaged}, first partitions the list
by using the first element as a pivot%
\footnote{We assume that the list is prepermuted to guarantee the expected
linear time behavior.}
and then recurs on one of the two resulting sides to find the desired element.
The side chosen is determined by the relationship of \texttt{k} to the size of
the first half \texttt{n}, which is returned by the \texttt{partition} function
along with the two sides themselves.

@@

An astute programmer might notice that much of the code in \texttt{quickselect}
does \emph{not} depend on the rank \texttt{k}; for example, the list is
partitioned before \texttt{k} is ever used. Moreover, while the recursive calls
to \texttt{qSelect} are guarded by a comparison to \texttt{k}, \texttt{k} does
not determine the arguments to those recursive calls---it only determines which
call is made. Thus, if we are willing to change the evaluation order of the
language (and evaluate under the \texttt{case} statement), it should be possible
to set things up so that \emph{all} the computation involving \texttt{l} is
performed before \emph{any} of the computation involving \texttt{k}.

We can formalize this intuition by writing quickselect in a \emph{staged}
language. In this paper, we choose \lang\ (\ref{sec:semantics}), a staged, typed
lambda calculus. This allows us to directly express the idea that the argument
\texttt{l} is known at the \emph{first} stage of the computation, while the
argument \texttt{k} is only known at the \emph{second}, and all
first-stage computation occurs before any second-stage computation.

We define a staged version of quickselect in \ref{fig:qs-staged}, writing
first-stage computations in red, and second-stage computations in blue. While
\texttt{qSelect} sends a \textrm{list} and \rmint\ to an \rmint,
\texttt{qsStaged} has a more precise type---it is a first-stage function which
takes a \textrm{list} and a $\fut\rmint$ (an $\rmint$ \emph{in the future}), and
returns a $\fut\rmint$.

\crem{Next three paragraphs are waiting on exact code for \ref{fig:qs-staged}.
Use three colors?}

The stage \bbone\ portions of \texttt{qsStaged} look like \texttt{qSelect}: any
term of non-circle type is available for immediate use. Indeed, we case on
\texttt{l} as usual, and in the \texttt{Cons} branch, \texttt{partition} it.

The next part of the original \texttt{qSelect} function depends on \texttt{k},
which has type $\fut\rmint$, indicating that it is only available to stage
\bbtwo\ computations. Since \texttt{qsStaged} itself produces a $\fut\rmint$,
the remainder of the function is stage \bbtwo\ code.

The $\next$ wraps a stage \bbtwo\ expression of type $\rmint$ (the \texttt{case}
expression) as a stage \bbone\ expression of type $\fut\rmint$ (the result type
of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
$\prev$ unwraps stage \bbone\ $\fut\rmint$s (the results of the recursive calls
to \texttt{qsStaged}) as stage \bbtwo\ $\rmint$s. In the \texttt{EQ} branch,
$\pause$ promotes an integer at stage \bbone\ (the head of \texttt{l}) directly
to an integer at stage \bbtwo.%
\footnote{$\pause$ can be implemented from our other language features, with
some effort. We provide it as a primitive to shorten examples. Furthermore, it
would be wise to extend $\pause$ to other base types, if we had them, and to
products and sums thereof.  This is related to the notion of {\em mobility} in
\cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}

@@

%The purpose of writing quickselect in \lang\ is that, because \texttt{qsStaged}
%typechecks, it is a valid way of dividing the program into two stages, where the
%stage \bbone\ computations can all be performed prior to the stage \bbtwo\
%computations. In \ref{sec:semantics} we provide an evaluation strategy for
%\lang\ which validates that claim.

%(We adopt the convention that the top-level context is stage \bbone.)

%pass around opaque stage \bbtwo\ computations at stage \bbone

@@

Intuitively, \texttt{l} determines the result of all calls to
\texttt{partition}, and \texttt{k} only determines which calls are made. So we
can preprocess \texttt{l} by recursively dividing it into halves smaller and
greater than the pivot---that is, building a binary search tree. Then, once we
have \texttt{k}, we can recur on this tree, choosing whichever branch has the
\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. Crucially,
because \texttt{partition} has no stage \bbtwo\ content, we need not perform any
of its work at stage \bbtwo!

\crem{Use $\bbonep$ to explain the above.}

@@

In this paper, we develop a splitting algorithm
(\ref{sec:splitting,sec:implementation}) which, for any term $e$ in \lang,
produces a pair of terms $e_1,e_2$ containing respectively all the first- and
second-stage content of $e$, such that $e_2~e_1$ is equivalent to $e$.
(Such a splitting is always possible because $e$ is well-typed in \lang.) In the
case of \texttt{qsStaged}, our splitting algorithm produces the algorithm
described above.

@@

>>>>

\begin{abstrsyn}

@@

\crem{
Notes from 1/16:
\begin{itemize}
\item We are modeling computations where some inputs arrive at stage 1 and
others arrive at stage 2. As a result, some subcomputations can be performed at
stage 1, and others at stage 2.
\item \emph{Statics}: M12 is a language with three worlds: 1M, 1P, and 2.
We model the computations described above as functions in world M, and their
results as values in 2.
\item \emph{Dynamics}: We describe how to perform all of the stage 1
subcomputations in an M12 program. Our semantics are mutually defined in terms
of how to reduce terms in each of the three worlds.
\end{itemize}
}

@@

Quickselect, which we define in \ref{fig:qs-unstaged}, first partitions the list
by using the first element as a pivot%
\footnote{We assume that the list is prepermuted to guarantee the expected
linear time behavior.}
and then recurs on one of the two resulting sides to find the desired element.
The side chosen is determined by the relationship of \texttt{k} to the size of
the first half \texttt{n}, which is returned by the \texttt{part}ition function
along with the two sides themselves.

@@

An astute programmer might notice that much of the code in \texttt{quickselect}
does \emph{not} depend on the rank \texttt{k}; for example, the list is
partitioned before \texttt{k} is ever used. Moreover, while the recursive calls
to \texttt{qSelect} are guarded by a comparison to \texttt{k}, \texttt{k} does
not determine the arguments to those recursive calls---it only determines which
call is made. Thus, if we are willing to change the evaluation order of the
language (and evaluate under the \texttt{case} statement), it should be possible
to force \emph{all} the computations involving \texttt{l} to be performed before
\emph{any} of the computations involving \texttt{k}.

We can formalize this intuition by writing quickselect in a \emph{staged}
language. In this paper, we choose \lang\ (\ref{sec:semantics}), a staged, typed
lambda calculus. This allows us to directly express the idea that the argument
\texttt{l} is known at the \emph{first} stage of the computation, while the
argument \texttt{k} is only known at the \emph{second}, and all
first-stage computations occur before any second-stage computations.

We define a staged version of quickselect in \ref{fig:qs-staged}, writing
first-stage computations in red, and second-stage computations in blue. While
\texttt{qSelect} sends a \textrm{list} and $\rmint$ to an $\rmint$,
\texttt{qsStaged} has a more precise type---it is a first-stage function which
takes a \textrm{list} (ignoring the $\curr$ for the moment) and a $\fut\rmint$
(an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.

Ignoring for the moment all \texttt{pure} annotations, the first-stage code in
\texttt{qsStaged} looks like \texttt{qSelect}: in the first stage, terms of
non-circle type are available for immediate use. Indeed, we case on the list as
usual, and in the \texttt{Cons} branch, \texttt{part}ition it.

The \texttt{case} expression in \texttt{qSelect} depends on \texttt{k}, whose
type $\fut\rmint$ indicates that it is only available to second-stage
computations. Since \texttt{qsStaged} itself produces a $\fut\rmint$, the
remainder of the function is second-stage code.

The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
expression) as a first-stage expression of type $\fut\rmint$ (the result type
of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
to \texttt{qsStaged}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
second-stage $\rmint$.%
\footnote{$\pause$ can be implemented from our other language features, with
some effort. We provide it as a primitive to shorten examples. Furthermore, it
would be wise to extend $\pause$ to other base types, if we had them, and to
products and sums thereof.  This is related to the notion of {\em mobility} in
\cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}

@@

Our type system, discussed in \ref{sec:semantics}, ensures that the staging
annotations are consistent, in the sense that computations marked as
first-stage cannot depend on ones marked as second-stage. The $\curr$ and
\texttt{pure} annotations further distinguish those first-stage computations
which do not contain any embedded second-stage code. (Code \emph{inside}
\texttt{pure}, like the \texttt{part}ition function, is guaranteed not to
contain second-stage code; other code, like \texttt{qsStaged}, may.) We will
discuss the importance of this additional distinction in \ref{sec:splitting}.

@@

Intuitively, \texttt{l} determines the result of all calls to
\texttt{part}ition, and \texttt{k} only determines which calls are made. So we
can preprocess \texttt{l} by recursively dividing it into halves smaller and
greater than the pivot---that is, building a binary search tree. Then, once we
have \texttt{k}, we can recur on this tree, choosing whichever branch has the
\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. And because
\texttt{part}ition contains no second-stage code, we can run it entirely in the
first stage!

@@

In this paper, we develop a splitting algorithm
(\ref{sec:splitting,sec:implementation}) which, given a program $e$ in \lang,
produces an equivalent pair of programs which correspond precisely to the two
stages of computation in $e$.
(Splitting is always possible because the staging annotations in $e$ are
consistent, because $e$ is well-typed in \lang.) In the case of
\texttt{qsStaged}, our splitting algorithm produces the algorithm described
above.

@@

\end{abstrsyn}
